---
title: "filter"
output: html_document
date: "2024-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
### setup ----------------------------------------------------------------------
source("config.R")
remotes::install_github("frvirard/signatR", auth_token = token)

# library ######################################################################
lib_lst <- c("dplyr",
             "purrr",
             "stringr")

invisible(purrr::walk(lib_lst, library, character.only = TRUE))
rm(lib_lst)

# functions ####################################################################

# variable #####################################################################
maf_dr <- file.path(data_dr, "maf")
```

The challenge was to find the right filtering to reduce the impact of FFPE induced mutations.
A previous work using VAFer was rather successful at extracting GA signature with MSA. Unfortunately, filtering was wrong (latter called bad filtering) and casting doute on the result.


## VAFer
We used VAFer with Jiri to find the more relevant setup. VAFer is providing a setting file. We started from sweet spot and increased the cutoff until reaching a mutation count around 3000.

```{r import mutect data}
param_file <- "2024-01-19_GWZ_20221025_MODARC_mutect_20221117.maf_settings.tsv"
maf_file <- "GWZ_20221025_MODARC_mutect_20221117.maf"

param <- vroom::vroom(file.path(data_dr, param_file), show_col_types = FALSE)
maf <- vroom::vroom(file.path(data_dr, "maf", maf_file), show_col_types = FALSE)
```

```{r filter}
output_file <- str_replace(maf_file, ".maf$", "_vafer.maf")

if(!file.exists(file.path(maf_dr, output_file))){
  dt <- param|>
  rename(s = sample, 
         cn = Cov_N,
         can = Cov_alt_N,
         ct = Cov_T,
         cat = Cov_alt_T)|>
  pmap_dfr(function(s, cn, 
                    can, ct, cat, vaf_cutoff, ...){
    message("oo processing ", s)
    maf|>
      filter(FILTER == "PASS")|>
      filter(avsnp150 == ".")|>
      filter(sample == s) |>
      filter(Cov_N >= cn) |>
      filter(Cov_alt_N <= can)|>
      filter(Cov_T >= ct)|>
      filter(Cov_alt_T >= cat)|>
      filter(VAF_T >= vaf_cutoff)
  })
  vroom::vroom_write(dt, file.path(data_dr, "maf", output_file))
}else{
  message("file detected, skipping ...")
}
```

Because filtering was based on spectra and total mutation count, I checked the mutation count after filtering.

```{r check mutation count}
vroom::vroom(file.path(data_dr, "maf",
                       "GWZ_20221025_MODARC_mutect_20221117_vafer.maf"),
             show_col_types = FALSE)|>
  mutate(mut_type = dplyr::case_when(Alt == "-" ~ "INDEL",
                                     Ref == "-" ~ "INDEL", 
                                     .default = "SNP"), .after = assembly_version)|>
  filter(mut_type != "INDEL")|>
  count(sample)
```

## old VAFer
The result wasn't the one expected after msa fitting. I created another maf with old settings to try understand the origin of the problem.
Note that there was an error in MORDAC_bak filtering Cov_T cutoff was set to 0 and not 10.

```{r vafer old}
param_file <- "2023-01-06_MODARC_newcall_mutect.maf_settings.tsv"
maf_file <- "GWZ_20221025_MODARC_mutect_20221117.maf"

param <- vroom::vroom(file.path(data_dr, param_file), show_col_types = FALSE)
maf <- vroom::vroom(file.path(data_dr, "maf", maf_file), show_col_types = FALSE)

output_file <- str_replace(maf_file, ".maf$", "_vafer_old.maf")

if(!file.exists(file.path(maf_dr, output_file))){
  dt <- param|>
  rename(s = sample, 
         cn = Cov_N,
         can = Cov_alt_N,
         ct = Cov_T,
         cat = Cov_alt_T)|>
  pmap_dfr(function(s, cn, 
                    can, ct, cat, vaf_cutoff, ...){
    message("oo processing ", s)
    maf|>
      filter(FILTER == "PASS")|>
      filter(avsnp150 == ".")|>
      filter(sample == s) |>
      filter(Cov_N >= cn) |>
      filter(Cov_alt_N <= can)|>
      filter(Cov_T >= ct)|>
      filter(Cov_alt_T >= cat)|>
      filter(VAF_T >= vaf_cutoff)
  })
  vroom::vroom_write(dt, file.path(data_dr, "maf", output_file))
}else{
  message("file detected, skipping ...")
}
```

## VAFer + cov_T >1
Jiri asked to test another filtering with >5 cov_mut for all samples and a VAF based on latest VAFer settings.

```{r vafer cov5}
param_file <- "2024-01-19_GWZ_20221025_MODARC_mutect_20221117.maf_settings.tsv"
maf_file <- "GWZ_20221025_MODARC_mutect_20221117.maf"

param <- vroom::vroom(file.path(data_dr, param_file), show_col_types = FALSE)
maf <- vroom::vroom(file.path(data_dr, "maf", maf_file), show_col_types = FALSE)

output_file <- str_replace(maf_file, ".maf$", "_vafer_cov5.maf")

if(!file.exists(file.path(maf_dr, output_file))){
  dt <- param|>
  rename(s = sample, 
         cn = Cov_N,
         can = Cov_alt_N,
         ct = Cov_T,
         cat = Cov_alt_T)|>
  pmap_dfr(function(s, cn, 
                    can, ct, cat, vaf_cutoff, ...){
    message("oo processing ", s)
    maf|>
      filter(FILTER == "PASS")|>
      filter(avsnp150 == ".")|>
      filter(sample == s) |>
      filter(Cov_N >= cn) |>
      filter(Cov_alt_N <= can)|>
      filter(Cov_T >= ct)|>
      filter(Cov_alt_T >= 5)|>
      filter(VAF_T >= vaf_cutoff)
  })
  vroom::vroom_write(dt, file.path(data_dr, "maf", output_file))
}else{
  message("file detected, skipping ...")
}
```
