---
title: "prepare MSA fitting"
output: html_document
date: "2024-01-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
### setup ----------------------------------------------------------------------
source("config.R")
remotes::install_github("frvirard/signatR", auth_token = token)

# library ######################################################################
lib_lst <- c("dplyr",
             "purrr",
             "stringr",
             "grid",
             "patchwork")

invisible(purrr::walk(lib_lst, library, character.only = TRUE))
rm(lib_lst)

# functions ####################################################################
# variable #####################################################################
sigpro_path <- file.path(data_dr, "sigprofiler")
```

I started by comparing denovo signature for all set
```{r compare denovo sign set, fig.width=20, fig.height= 10}
set_lst <- dir(file.path(sigpro_path, "sign"))|>
  str_subset("strelka2|KIRC-US|RECA-EU", negate = TRUE)|>
  set_names()

sign_profile <- map(set_lst, signatR::sigpro_import_sign, type = "profile",
                                        n_sign = 4, 
                                        analysis = "denovo", 
                                        context = "SBS288")
sign_profile <- sign_profile[c(2,3,4,1)]

for(s in 2:length(sign_profile)){
    sign_profile[[s]] <- signatR::sign_reorder(dt_sign = sign_profile[[s]], 
                                      dt_ref = sign_profile[[s-1]])
}

sign_plot <- imap(sign_profile, function(sign_act, set_name){
  plot <- signatR::plot_spectra(sign_act, context = "SBS192", render = FALSE)
  title <- textGrob(set_name)
  
  wrap_elements(title)/wrap_elements(plot)+plot_layout(heights = c(0.1, 0.9))
})
  
wrap_plots(sign_plot)+plot_layout(ncol = 4)
```

## VAFer
We decided with Jiri to do a first run using solution 4 of Vafer filtering.
SBS288B was considered as noise and was added to SBS40, 5, 1 and 8 with SBS_GA for MSA fitting.

```{r}
sign_dt <- signatR::sigpro_import_sign("GWZ_20221025_MODARC_mutect_20221117_vafer",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = 4, 
                            analysis = "denovo")

signatR::plot_spectra(sign_dt, context = "SBS192")

noise_sign <- sign_dt|>
  select(MutationType, SBS288B)|>
  filter(!str_detect(MutationType, "N"))|>
  mutate(Strand = str_sub(MutationType, 1,1),
         Type = str_sub(MutationType, 5,7),
         Subtype = paste0(str_sub(MutationType, 3,3),
                          str_sub(MutationType, 5,5),
                          str_sub(MutationType, 9,9)))|>
      select(Strand, Type, Subtype, SBS_FFPE = SBS288B)|>
  mutate(SBS_FFPE = SBS_FFPE/sum(SBS_FFPE))

sbs_GA <- vroom::vroom(file.path(file_dr, "MEFh_GA_WX_192.csv"), show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS_GA = MEFh_GA_WX_192)

custom_cosmic <- vroom::vroom(file.path(file_dr, "IARC_custom_SBS_192_signatures.csv"), 
                              show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS1, SBS5, SBS40, SBS8)

custom_cosmic|>
  left_join(sbs_GA)|>
  left_join(noise_sign)|>
  vroom::vroom_write(file.path(file_dr, "custom_1_5_40_8_FFPE_GA_SBS_192_signatures.csv"), 
                      delim = ",")
```

## VAFer old
with a VAFer filtering less stringent (old version), signature were nearly identical.
FFPE noise was signature SBS288C.

```{r}
sign_dt <- signatR::sigpro_import_sign("GWZ_20221025_MODARC_mutect_20221117_vafer_old",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = 4, 
                            analysis = "denovo")

signatR::plot_spectra(sign_dt, context = "SBS192")
```

```{r}
ffpe_old <- signatR::sigpro_import_sign("GWZ_20221025_MODARC_mutect_20221117_vafer_old",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = 4, 
                            analysis = "denovo")|>
  select(MutationType, ffpe_old = SBS288C)

ffpe_new <-  signatR::sigpro_import_sign("GWZ_20221025_MODARC_mutect_20221117_vafer",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = 4, 
                            analysis = "denovo")|>
  select(MutationType, ffpe_new = SBS288B)

ffpe_comp <- ffpe_new|>
  left_join(ffpe_old)

signatR::plot_spectra(ffpe_comp, context = "SBS192")

lsa::cosine(ffpe_comp$ffpe_new, ffpe_comp$ffpe_old)
```

I also prepared a signature set from previous bad filtering to test reproducibility. of GA fitting.

```{r}
FFPE_sign <- "SBS288C"

noise_sign <- sign_dt|>
  select(MutationType, all_of(FFPE_sign))|>
  filter(!str_detect(MutationType, "N"))|>
  mutate(Strand = str_sub(MutationType, 1,1),
         Type = str_sub(MutationType, 5,7),
         Subtype = paste0(str_sub(MutationType, 3,3),
                          str_sub(MutationType, 5,5),
                          str_sub(MutationType, 9,9)))|>
      select(Strand, Type, Subtype, SBS_FFPE = FFPE_sign)|>
  mutate(SBS_FFPE = SBS_FFPE/sum(SBS_FFPE))

sbs_GA <- vroom::vroom(file.path(file_dr, "MEFh_GA_WX_192.csv"), show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS_GA = MEFh_GA_WX_192)

custom_cosmic <- vroom::vroom(file.path(file_dr, "IARC_custom_SBS_192_signatures.csv"), 
                              show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS1, SBS5, SBS40, SBS8)

custom_cosmic|>
  left_join(sbs_GA)|>
  left_join(noise_sign)|>
  vroom::vroom_write(file.path(file_dr, "custom_1_5_40_8_FFPE_old_GA_SBS_192_signatures.csv"), 
                      delim = ",")
```

## bad filtered

```{r}
sign_dt <- signatR::sigpro_import_sign("MODARC_mutect_20221117_VAFer_bad_filtered",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = "suggested", 
                            analysis = "denovo")
signatR::plot_spectra(sign_dt, context = "SBS192")
```

```{r fig.width=10}
dt_sign <- signatR::sigpro_import_sign("MODARC_mutect_20221117_VAFer_bad_filtered",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = "suggested", 
                            analysis = "denovo")

dt_ref <- signatR::sigpro_import_sign("GWZ_20221025_MODARC_mutect_20221117_vafer",
                            type = "profile", 
                            context = "SBS288",
                            n_sign = 4, 
                            analysis = "denovo")

dt_sign <- sign_reorder(dt_sign, dt_ref)

ref <- signatR::plot_spectra(dt_ref, context = "SBS192", render = FALSE)
left <- signatR::plot_spectra(dt_sign, context = "SBS192", render = FALSE)

wrap_plots(ref)+wrap_plots(left)

```
```{r}
FFPE_sign <- "SBS288A"

noise_sign <- dt_sign|>
  select(MutationType, all_of(FFPE_sign))|>
  filter(!str_detect(MutationType, "N"))|>
  mutate(Strand = str_sub(MutationType, 1,1),
         Type = str_sub(MutationType, 5,7),
         Subtype = paste0(str_sub(MutationType, 3,3),
                          str_sub(MutationType, 5,5),
                          str_sub(MutationType, 9,9)))|>
      select(Strand, Type, Subtype, SBS_FFPE = FFPE_sign)|>
  mutate(SBS_FFPE = SBS_FFPE/sum(SBS_FFPE))

sbs_GA <- vroom::vroom(file.path(file_dr, "MEFh_GA_WX_192.csv"), show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS_GA = MEFh_GA_WX_192)

custom_cosmic <- vroom::vroom(file.path(file_dr, "IARC_custom_SBS_192_signatures.csv"), 
                              show_col_types = FALSE)|>
    select(Strand, Type, Subtype, SBS1, SBS5, SBS40, SBS8)

custom_cosmic|>
  left_join(sbs_GA)|>
  left_join(noise_sign)|>
  vroom::vroom_write(file.path(file_dr, "custom_1_5_40_8_FFPE_bad_filtered_GA_SBS_192_signatures.csv"), 
                      delim = ",")
```

